Making ice sheet models \\ scale properly

### 2023-01-03 talk/slides.tex
progress slides

\newcommand{\bxi}{\bm{\xi}}

@@

\titlegraphic{\hfill \includegraphics[width=0.2\textwidth]{images/uafbw.png}}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon!
    \end{itemize}
\item ice sheet geometry and velocity evolve \emph{in contact with the climate}
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt and calving when floating in the ocean (\emph{ice shelves})
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming performance analysis, I will be ignoring quite a bit of detail!
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \end{itemize}
\item now, a simple definition of \emph{ice sheet model}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a map
\end{definition} 
\item two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$a(t,x) = \begin{pmatrix}
\text{balance of solid precipitation} \\
\text{\, vs melt \& runoff at surfaces}
\end{pmatrix}$$

@@

    \item[$\circ$] \emph{bed elevation} $b(x)$ [m]
    \end{itemize}
\item one output:
    \begin{itemize}
    \item[$\circ$] \emph{upper surface elevation} $s(t,x)$ [m]
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/coverfig.png}

@@

\begin{itemize}
\item but this is a mathematics talk in fact
\item FIXME conservation laws; conserve momentum standard; conserve mass means both conserve mass in interior (incompressible) and at surface (SKE)
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item FIXME RETURN TIME DERIV
\item nonlinear complementarity problem coupled to Stokes:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega$} \\
s - \Delta t\, \bu|_s \cdot \bn_s - (\sold + \Delta t\, a )&\ge 0 && \text{''} \\
(s - b) (s - \Delta t\, \bu|_s \cdot \bn_s - (\sold + \Delta t\, a)) &= 0 && \text{''} \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\bu &= \bzero && \text{on $\Gamma_0$} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$}
\end{align*}

    \begin{itemize}
    \item with regularized Glen-law effective viscosity ($\text{p}=\frac{1}{\nn}+1$):
      $$\nu_\eps(D\bu) = \frac{\Gamma}{2} \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\section{scalable implicit steps?}

@@

\begin{itemize}
\item x
\item y
\end{itemize}
\end{frame}

@@

\begin{itemize}
\footnotesize
\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
\item x
\end{itemize}
\end{frame}

>>>>

\newcommand{\btau}{\bm{\tau}}
\newcommand{\bxi}{\bm{\xi}}

@@

\titlegraphic{\hfill \includegraphics[width=0.15\textwidth]{images/uafbw.png}}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon!
    \end{itemize}
\item ice sheet geometry and velocity evolve \emph{in contact with the climate}
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] melt and calving when floating in the ocean (\emph{ice shelves})
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming performance analysis, I will be ignoring quite a bit of physics!
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \end{itemize}
\item now, a simple definition of ``ice sheet model'' \dots
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\begin{definition}
an \alert{ice sheet model} is a \underline{map} which simulates an ice sheet in a climate
\end{definition} 
\item two inputs:
    \begin{itemize}
    \item[$\circ$] \emph{surface mass balance}
$$a(t,x) = \begin{pmatrix}
\text{balance of solid precipitation} \\
\text{\, vs melt \& runoff at surfaces}
\end{pmatrix}$$

@@

    \item[$\circ$] \emph{bed elevation} $b(x)$
    \end{itemize}
\item two outputs:
    \begin{itemize}
    \item[$\circ$] \emph{upper surface elevation} $s(t,x)$
    \item[$\circ$] \emph{ice velocity} $\bu(t,x)$
    \end{itemize}
\item {\small $\begin{pmatrix} \text{climate} \\ \text{topography} \end{pmatrix} \to \begin{pmatrix} \text{geometry} \\ \text{velocity} \end{pmatrix}$}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-velocity.png}

@@

\begin{itemize}
\item as this is a mathematics talk, more or less, some notation \dots
\item FIXME conservation laws; conserve momentum standard; conserve mass means both conserve mass in interior (incompressible) and at surface (SKE)
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item nonlinear complementarity problem coupled to Stokes:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\bu\cdot \bn = 0 \quad \& \quad \btau_b - \bbf(\bu) &= \bzero && \text{on $\Gamma_b(t) \subset \partial \Lambda(t)$} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t) \subset \partial \Lambda(t)$}
\end{align*}

    \begin{itemize}
    \item viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\section{scalable implicit steps by multigrid?}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item current-technology ice sheet models mostly use \alert{explicit time stepping}, \alert{non-optimal stress-balance solvers}, and shallow assumptions
\item \alert{coming soon} from current research:
   \begin{itemize}
   \item[$\circ$] scalable Stokes solvers (Isaac et al.~2015)
   \item[$\circ$] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[$\circ$] ML emulators (Jouvet et al.~2021)
   \end{itemize}
\item scalable solvers for the implicit-step, coupled problem for ice sheet geometry requires advances in \alert{multilevel solutions of non-local variational inequalities}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\footnotesize
\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
\item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015). \emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
\item G.~Jouvet, G.~Cordonnier, B.~Kim, M.~L\"uthi, A.~Vieli, A.~Aschwanden (2021). \emph{Deep learning speeds up ice flow modelling by several orders of magnitude}, J.~Glaciol.~\href{https://doi.org/10.1017/jog.2021.120}{10.1017/jog.2021.120}
\item A.~L{\"o}fgren, J. Ahlkrona, \& C. Helanow (2022). \emph{Increasing stable time-step sizes of the free-surface problem arising in ice-sheet simulations}, J.~Comput.~Phys.~X 16, \href{https://doi.org/10.1016/j.jcpx.2022.100114}{10.1016/j.jcpx.2022.100114}
\end{itemize}
\end{frame}
